---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# metaboprep2

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/nicksunderland/metaboprep2/graph/badge.svg)](https://app.codecov.io/gh/nicksunderland/metaboprep2)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`metaboprep2` is an R package designed for processing and managing metabolomic data efficiently. It provides structured storage and access to metabolomic datasets, including sample metadata, feature metadata, and metabolite intensity data. Whilst the class structure is new, the underlying routines are based on the excellent work from the authors of [metaboprep](https://github.com/MRCIEU/metaboprep). Until this project is stable and tested it is recommended you use the original `metaboprep` pipeline. 

## Installation

You can install the development version of metaboprep2 from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("pak")
pak::pak("nicksunderland/metaboprep2")

# or 
remotes::install_github("nicksunderland/metaboprep2")
```

## Getting Started

Please see the dedicated [vignettes](https://nicksunderland.github.io/metaboprep2/) for common use scenarios:  

* Importing data
* Running the quality control pipeline  
* Generating a QC report  
* Exporting data  


