---
title: "Using the Metabolites S7 Class"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Metabolites S7 Class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `Metabolites` class is part of the `metaboprep2` package and is designed to store and manage metabolomic data efficiently. It provides structured access to sample metadata, feature metadata, and metabolite intensity data.

## Installation and Loading

Ensure that you have installed and loaded the `metaboprep2` package before using the `Metabolites` class:

```{r}
# Install (if necessary)
# install.packages("metaboprep2")

# Load package
library(metaboprep2)
```

## Creating a Metabolites Object

A `Metabolites` object is created using the constructor function. It requires a file path to the data and a project name. You will need to specify the format of the raw data file - the available options can be found using
the function `available_data_formats()`

```{r}
available_data_formats()
```

```{r}
# Example Metabolon (format 1) raw data file
filepath <- system.file("extdata", "metabolon_v1_example.xlsx", package = "metaboprep2")

# Create a Metabolites object
metabolites <- Metabolites(project_name = "ExampleProject", 
                           filepath     = filepath,
                           format       = "metabolon_v1")

# Import the data 
metabolites <- import_data(metabolites)

# View
metabolites
```

## Metabolites Object Structure

A `Metabolites` object consists of three main data components:

- **samples**: A `data.table` containing metadata about the samples (e.g., sample IDs, collection details).
- **features**: A `data.table` describing the metabolites, including feature names and additional annotations.
- **data**: A multi-dimensional matrix where:
  - Rows represent individual samples.
  - Columns correspond to different metabolites.
  - The third dimension (depth) represents various stages of data processing (e.g., raw, scaled, post-QC).
- **exclusions**: A multi-dimensional sparse character matrix in which each layer is a mask for the corresponding layer in the @data matrix slot. The cells contain a comma separated string of exclusion codes (see the Quality Control article).
  
The `Metabolites` object also holds parameters for other analyses, such as quality control routines. There are 
default values, however these can either be changed during object creation, or later using the `@` operator to 
access the object's slots. 


## Accessing Data

There are two primary ways to access data stored in a `Metabolites` object:

1. **Using the `@` operator** to directly access object slots.
2. **Using getter functions** like `get_data()` for structured data extraction.

### Accessing Data Using `@` Operator

#### Sample Metadata
```{r}
head(metabolites@samples)
```

#### Feature Metadata
```{r}
head(metabolites@features)
```

#### Raw Metabolite Data
```{r}
str(metabolites@data)
```

### Accessing Data Using Getter Functions

The `get_data()` function allows retrieving metabolite data with optional filtering.

#### Retrieve All Data of a Given Type
```{r}
original_data <- get_data(metabolites, layer = "raw")
original_data[1:5, 1:5]
```

#### Extract Specific Metabolites and Samples
```{r}
subset_data <- get_data(metabolites, 
                        layer = "raw", 
                        feature_ids = c("compid_123", "compid_124"), 
                        sample_ids = c("ind1", "ind2"))
subset_data
```


## Exporting Data

The `write_data()` method allows saving the structured data to a specified directory.

```{r, eval=FALSE}
write_data(metabolites, directory = "output_directory")
```

## Summary

The `Metabolites` class provides an organized way to handle metabolomics data, making it easier to access, manipulate, and export datasets for further analysis. See the other vignettes and functions for quality control reporting and analysis routines.

---
