---
title: "Metabolite Quality Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metabolite Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## **Introduction**  

Metabolomics data often contain missing values, outliers, and features that require filtering before downstream analysis. The `metabolite_qc()` function provides an efficient and automated way to perform **quality control (QC)** on a `Metabolites` object.  

This vignette walks through the key parameters and steps for **performing QC** using this function.  

---

## **Creating a Metabolites Object**  

Before performing QC, you need a `Metabolites` object, which contains metabolomics data along with metadata.  

```{r setup}
library(metaboprep2)

# Example Metabolon (format 1) raw data file
filepath <- system.file("extdata", "metabolon_v1_example.xlsx", package = "metaboprep2")

# Create a Metabolites object with the desired parameters
metabolites <- Metabolites(project_name              = "ExampleProject", 
                           filepath                  = filepath,
                           format                    = "metabolon_v1",   
                           feature_missingness       = 0.2,
                           sample_missingness        = 0.2,
                           total_peak_area_sd        = 5,
                           outlier_udist             = 5,
                           outlier_treatment         = "leave_be",
                           winsorize_quantile        = 1,
                           tree_cut_height           = 0.5,
                           pc_outlier_sd             = 5,
                           derived_var_exclusion     = TRUE,
                           xenobiotics_var_exclusion = TRUE)

# Import the data 
metabolites <- import_data(metabolites)
```

## **Performing Metabolite Quality Control**

The metabolite_qc() function applies multiple QC filters based on missingness, outlier detection, PCA-based QC. This generates another layer to the data matrices named `post_qc`.

```{r}
metabolites <- metabolite_qc(metabolites, source = "raw", destination = "post_qc")
```

### Accessing QC'd data

There are two primary ways to access data stored in a `Metabolites` object:

1. **Using the `@` operator** to directly access object slots.
2. **Using getter functions** like `get_data()` for structured data extraction.

#### Accessing Data Using `@` Operator
```{r}
metabolites@data[, , "post_qc"][1:5,1:5]
```

#### Accessing Data Using Getter Functions
```{r}
get_data(metabolites, type="post_qc", apply_exclusions = FALSE)[1:5,1:5]
```

#### Accessing Only Data Passing QC
```{r}
get_data(metabolites, type="post_qc", apply_exclusions = TRUE)[1:5,1:5]
```

### Exclusions
The underlying data still has all respresentations of the samples and features, however, the QC process builds a matrix of exclusion reasons.

#### Accessing the Exclusions Matrix
```{r}
metabolites@exclusions[, , "post_qc"][1:5,1:5]
```
#### Accessing the Exclusions Codes
```{r}
get_exclusion_codes(verbose=TRUE)
```

### Figures
Currently the feature tree object is stored in the metabolites object. 

More DataViz to come....
```{r, fig.width=14, fig.height=8, out.width="100%"}
plot(metabolites@feature_tree[["post_qc"]])
```

