---
title: "Metabolite Quality Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metabolite Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The quality control pipeline can be implemented with the following lines of code, namely the `metabolite_qc()` function. You must specify the QC parameters in the slots of the `Metabolomics` object, this can either be done at object instantiation, or by later setting the slots.  

This vignette walks through the key parameters and steps for **performing QC** using this function.  

---

## **Creating a Metabolites Object**  

Before performing QC, you need a `Metabolites` object, which contains metabolomics data along with metadata.  

```{r setup}
library(metaboprep2)

# Example Metabolon (format 1) raw data file
filepath <- system.file("extdata", "metabolon_v1_example.xlsx", package = "metaboprep2")

# Create a Metabolites object with the desired parameters
metabolites <- Metabolites(project_name              = "ExampleProject", 
                           filepath                  = filepath,
                           format                    = "metabolon_v1",   
                           feature_missingness       = 0.2,
                           sample_missingness        = 0.2,
                           total_peak_area_sd        = 5,
                           outlier_udist             = 5,
                           outlier_treatment         = "leave_be",
                           winsorize_quantile        = 1,
                           tree_cut_height           = 0.5,
                           pc_outlier_sd             = 5,
                           derived_var_exclusion     = TRUE,
                           xenobiotics_var_exclusion = TRUE)

# Import the data 
metabolites <- import_data(metabolites)
```

## **Performing Metabolite Quality Control**

The `metabolite_qc()` function applies multiple QC filters based on missingness, outlier detection, PCA-based QC. This generates another layer to the data matrices named `post_qc`. And intermediate `qcing` layer is also generated.

```{r}
metabolites <- metabolite_qc(metabolites, source = "raw")
```

### Figures
Currently the feature tree object is stored in the metabolites object `@feature_tree` slot. 

```{r, fig.width=14, fig.height=8, out.width="100%"}
plot(get_feature_tree(metabolites, layer="post_qc"))
```

## **Report**
Please see the dedicated vignette on generating a QC report (html or pdf).
